---
title: "Storm Data Analysis - Course Project 2"
author: "Ashley Machado"
date: "January 26, 2016"
output: html_document
---

This project will involve analysis of Storm Data from U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database
As part of the analysis, we will try to answer 2 questions: 
1. Identify the most harmful weather events,across United States, to the population's health.
                                      and 
2. Identify the weather events which have the greatest economic consequences. 

###Data pre-processing section:

Loading relevant libraries
```{r, echo=FALSE, message=FALSE}
if(!require(dplyr)) {
    install.packages("dplyr")
    library(dplyr)  
  }
  if(!require(ggplot2)) {
    install.packages("ggplot2")
    library(ggplot2)  
  }
  if(!require(lattice)) {
    install.packages("lattice")
    library(lattice)  
  }
```

The storm data, which is in bz2 zipfile format will be read into a R dataframe.

```{r,  echo=TRUE, message=FALSE, cache=TRUE}
stormDF <- read.csv(bzfile("repdata-data-StormData.csv.bz2"))
```

To answer both the questions, we will subset the dataset to get only the weather event type(EVTYPE), fatalities(FATALITIES), injuries(INJURIES),Property Damage(PROPDMG),Property damage amount's exponent(PROPDMGEXP), Crop Damage(CROPDMG) and Crop damage amount's exponent(CROPDMGEXP) columns(variables).

To consolidate the harm done to the population, We will add up the fatalities(FATALITIES) and injuries(INJURIES) to create a new column(variable) - HARM - for the weather event types(EVTYPE).

The property and crop damage amounts have exponent values columns -PROPDMGEXP and CROPDMGEXP. 
And the exponent values are represented by alphabetical characters - "K" for thousands, "M" for millions, and "B" for billions. 
We will convert these alphabetical characters to their exponential values and then multiply them to their corresponding damage amounts. 
Any alphabet character other than - "K", "M", "B" - will be coverted to Zero(0) value, which will result in a Zero(0) damage amount.

```{r, echo=TRUE, message=FALSE}

subsetDF <- subset(stormDF, select = c("EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP"))

subsetDF$HARM <- subsetDF$FATALITIES+subsetDF$INJURIES

subsetDF <- subsetDF %>% mutate(
  PROPDMGEXP = ifelse (PROPDMGEXP == "K",1*10^3, 
              ifelse (PROPDMGEXP == "M",1*10^6,
               ifelse (PROPDMGEXP == "B",1*10^9,0))) ,
               
  CROPDMGEXP = ifelse (CROPDMGEXP == "K",1*10^3, 
              ifelse   (CROPDMGEXP == "M",1*10^6,
               ifelse   (CROPDMGEXP == "B",1*10^9,0))) 
  )

subsetDF$PROPDMG <- subsetDF$PROPDMG*subsetDF$PROPDMGEXP
subsetDF$CROPDMG <- subsetDF$CROPDMG*subsetDF$CROPDMGEXP

uniqueEV <- length(unique(subsetDF$EVTYPE))

```

###Results

To answer the first question of identifying  weather events most harmful to human population, we will generate 3 different views for the top 10 harmful weather events.

The first view will be the combined Fatalities and Injuries(count) because of Weather events. 
```{r, echo=TRUE, fig.height= 4, fig.width= 9}
subsetCombHarm <- subsetDF %>% group_by(EVTYPE) %>% summarise(CombinedImpact = sum(HARM)) %>% arrange(desc(CombinedImpact))
subsetCombHarm <- as.data.frame(subsetCombHarm)
subsetCombHarm <- subsetCombHarm[order(subsetCombHarm[,2], decreasing = TRUE),]
ggplot(data = subsetCombHarm[1:10,], aes(x=EVTYPE, 
    y=reorder(CombinedImpact, -table(CombinedImpact)[CombinedImpact])
    , fill=factor(CombinedImpact))) +      geom_bar(position="dodge",stat="identity") + coord_flip() + 
  ggtitle("Top 10 Weather events which led to Fatalities and Injuries")+ 
  ylab("Combined Fatalities and Injuries count")+
  xlab("Top Ten Weather events")
```

The second view will be the Fatality(count) because of weather events.
```{r, echo=TRUE, fig.height= 4, fig.width= 9, fig.align="RIGHT"}
subsetFatalHarm <- subsetDF %>% group_by(EVTYPE) %>% summarise(FatalHarm = sum(FATALITIES)) %>% arrange(desc(FatalHarm))
subsetFatalHarm <- as.data.frame(subsetFatalHarm)
subsetFatalHarm <- subsetFatalHarm[order(subsetFatalHarm[,2], decreasing = TRUE),]
ggplot(data = subsetFatalHarm[1:10,], aes(x=EVTYPE, 
    y=reorder(FatalHarm, -table(FatalHarm)[FatalHarm])
    , fill=factor(FatalHarm))) +      geom_bar(position="dodge",stat="identity") + coord_flip() + 
  ggtitle("Top 10 Weather events which led to Fatalities")+ 
  ylab("Fatality count")+
  xlab("Top Ten Weather events")

```

The Third view will be the Injured(count) because of weather events.
```{r, echo=TRUE, fig.height= 4, fig.width= 9}

subsetInjuryHarm <- subsetDF %>% group_by(EVTYPE) %>% summarise(InjuryHarm = sum(INJURIES)) %>% arrange(desc(InjuryHarm))
subsetInjuryHarm <- as.data.frame(subsetInjuryHarm)
subsetInjuryHarm <- subsetInjuryHarm[order(subsetInjuryHarm[,2], decreasing = TRUE),]
ggplot(data = subsetInjuryHarm[1:10,], aes(x=EVTYPE, 
    y=reorder(InjuryHarm, -table(InjuryHarm)[InjuryHarm])
    , fill=factor(InjuryHarm))) +      geom_bar(position="dodge",stat="identity") + coord_flip() + 
  ggtitle("Top 10 Weather events which led to Injuries")+ 
  ylab("Injured count")+
  xlab("Top Ten Weather events")

```


To identify the weather events which have the greatest economic consequences. we will generate 3 different views for top 10 weather events.

The First view will be the property damage cost because of the weather events.

```{r, echo=TRUE, fig.height= 4, fig.width= 9}
options(scipen = 10)

PropDmgDF <- subsetDF %>% group_by(EVTYPE) %>% summarise(PropertyDamage = sum(PROPDMG)) %>% arrange(desc(PropertyDamage))
PropDmgDF <- as.data.frame(PropDmgDF)
PropDmgDF <- PropDmgDF[order(PropDmgDF[,2], decreasing = TRUE),]
ggplot(data = PropDmgDF[1:10,], aes(x=EVTYPE, 
    y=reorder(PropertyDamage, -table(PropertyDamage)[PropertyDamage])
    , fill=factor(PropertyDamage))) +      geom_bar(position="dodge",stat="identity") + coord_flip() + 
  ggtitle("Top 10 Weather events which led to Property Damage")+ 
  ylab("Property Damage cost")+
  xlab("Top Ten Weather events")

```


The Second view will be the Crop damage cost because of the weather events.

```{r, echo=TRUE, fig.height= 4, fig.width= 9}

CropDmgDF <- subsetDF %>% group_by(EVTYPE) %>% summarise(CropDamage = sum(CROPDMG)) %>% arrange(desc(CropDamage))
CropDmgDF <- as.data.frame(CropDmgDF)
CropDmgDF <- CropDmgDF[order(CropDmgDF[,2], decreasing = TRUE),]
ggplot(data = CropDmgDF[1:10,], aes(x=EVTYPE, 
    y=reorder(CropDamage, -table(CropDamage)[CropDamage])
    , fill=factor(CropDamage))) +      geom_bar(position="dodge",stat="identity") + coord_flip() + 
  ggtitle("Top 10 Weather events which led to Crop Damage")+ 
  ylab("Crop Damage cost")+
  xlab("Top Ten Weather events")

```

The Third view will be the Combined Property and Crop damage cost because of the weather events.

```{r, echo=TRUE, fig.height= 4, fig.width= 9}

subsetDF$CombDmg <- subsetDF$PROPDMG+subsetDF$CROPDMG

CombDmgDF <- subsetDF %>% group_by(EVTYPE) %>% summarise(CombinedDamage = sum(CombDmg)) %>% arrange(desc(CombinedDamage))
CombDmgDF <- as.data.frame(CombDmgDF)
CombDmgDF <- CombDmgDF[order(CombDmgDF[,2], decreasing = TRUE),]
ggplot(data = CombDmgDF[1:10,], aes(x=EVTYPE, 
    y=reorder(CombinedDamage, -table(CombinedDamage)[CombinedDamage])
    , fill=factor(CombinedDamage))) +      geom_bar(position="dodge",stat="identity") + coord_flip() + 
  ggtitle("Top 10 Weather events which led to Combined Property and Crop Damage")+ 
  ylab("Combined Property and Crop Damaged Cost")+
  xlab("Top Ten Weather events")

```





